The Design and Implementation of the Synchronous Language \CEU

### 2016-04-04 paper.tex
...

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with support for a simpler 
semantics and more fine-grained execution control.
%
\CEU employs an event-triggered notion of time which enables compile-time 
analysis for simultaneous reactions to produce deterministic and 
concurrency-safe programs.
%
We discuss the particularities of our design in comparison to Esterel, such as 
stack-based internal events, temporal analysis for concurrent statements, safe 
integration with \C, and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

\subsection{Summary}

\CEU aims to offer a simpler semantics than Esterel with more determinism and 
fine-grained control over the program execution.
%
The list that follows summarizes the contributions of our design in this 
direction:
%
\begin{itemize}
%
\item Event-triggered notion of time bound to the semantics of the language.
Event-driven programming is popular in many domains, such as server and GUI 
development.
We believe that programmers are more familiar with dealing with events in 
isolation, which simplifies the reasoning about concurrency.
In addition, the uniqueness of external events is a prerequisite for the 
temporal analysis of \CEU.
%
\item Deterministic intra-reaction execution and communication.
Determinism in \CEU is ``all-inclusive'' and does not depend on additional 
levels of static analysis.
It encompasses the whole language, including memory accesses, system calls, and 
stack-based internal events.
Programmers can always figure out which statement executes next, making runtime 
analysis and debugging easier.
%
\item Static temporal analysis for concurrent statements.
Although execution is deterministic, the compiler of \CEU still advises about 
suspicious statements that can react concurrently to the same event, forcing 
the to serialize them accordingly.
%
\item Safe integration with \C.
\CEU enforces the programmer to deal with pointers representing resources 
through finalization clauses.
When dealing with concurrent system calls, programmers can provide the compiler 
with annotations to remove false positives from the temporal analysis or to 
allow non commutative concurrent behavior.
%
\item First-class timers with dedicated syntax and automatic synchronization.
Given the omnipresence of timers in embedded systems, a dedicated syntax can 
simplify the development and readability of programs.
Furthermore, automatic synchronization releases the programmer from the burden 
of adjusting timers in sequence and in parallel.
%
\end{itemize}

@@

\CEU has a strong influence from Esterel but differ in the most fundamental 
aspect of notion of time.
%
In \CEU, instead of clock ticks, the occurrence of a single external event 
defines an unit of time.
%
The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
We consider that this approach is more familiar to programmers and simplifies 
the reasoning about concurrency.
%
As far as we know, \CEU is unique in this notion of time encrusted in the core 
of the language, which is a prerequisite for the temporal analysis that enables 
safe shared-memory concurrency.

Another unique aspect of \CEU is the distinction between external and internal 
events.
Internal events define stack-based micro reactions within external reactions, 
providing more fine-grained control for intra-reaction execution.

Like \CEU, many synchronous languages rely on deterministic scheduling to 
preserve intra-reaction determinism (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%
%It is the responsibility of the programmer to specify the execution order for 
%threads, based on either explicit priorities, or source code lexical order 
%(similar to \CEU).
%
\CEU goes one step further and performs a temporal analysis to detect 
concurrent trails that actually rely on deterministic scheduling, i.e., the 
observable behavior of the program changes if they are reordered.
%
%These languages have a tick-based notion of time similar to Esterel, which 
%prevents the temporal analysis of \CEU.
%
\CEU also guarantees deterministic behavior for composition of timers by 
adjusting residual awaking times from inaccurate system clocks.

Regarding resource management, Esterel supports a finalization mechanism to 
unconditionally execute a series of statements on abortion.
%
\CEU also tracks pointers representing resources that cross \C boundaries and
enforce the programmer to provide corresponding finalizers.

@@

We present the design and implementation of \CEU, a synchronous reactive 
language inspired by Esterel with support for a simpler semantics and more 
fine-grained control for intra-reaction execution.

@@

\CEU is a practical language with expressive control constructs, such as 
lexically scoped parallel compositions, convenient first-class timers, and a 
unique stack-based internal signaling mechanism.
%
Programs interoperate seamlessly with \C, and can take advantage of existing 
libraries, lowering the entry barrier for adoption.
%
\CEU has an open source implementation and bindings for \emph{TinyOS}, 
\emph{Arduino}, and the \emph{SDL} graphical library.%
\footnote{Website of \CEU: \url{http://www.ceu-lang.org/}}

For the past three years, we have been teaching \CEU for undergraduate and 
graduate students in research projects and two hands-on courses on 
\emph{distributed systems} and \emph{reactive programming}.
%, in which the studnets designing protocols, embedded systems, and graphical 
%applications.
%
Our experience shows that students take advantage of the sequential-imperative 
style of \CEU and can implement non-trivial concurrent applications in a few 
weeks.

>>>>

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with simpler semantics and 
more fine-grained control over program execution.
%
\CEU adopts an event-triggered notion of time which enables compile-time 
analysis for simultaneous reactions to produce deterministic and 
concurrency-safe programs.
%
Using Esterel as a base, we discuss the particularities of our design, such as 
stack-based internal events, temporal analysis for concurrent statements, safe 
integration with \C, and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

\subsection{Summary}
\CEU aims to offer a simpler semantics than Esterel with more determinism and 
fine-grained control over program execution.
%
The list that follows summarizes the contributions of our design in this 
direction:
%
\begin{itemize}
%
\item Event-triggered notion of time bound to the semantics of the language.
Event-driven programming is popular in many domains, such as server and GUI 
development.
We believe that programmers are more familiar with dealing with events in 
isolation, which simplifies the reasoning about concurrency.
In addition, the uniqueness of external events is a prerequisite for the 
temporal analysis of \CEU.
%
\item Deterministic intra-reaction execution and communication.
Determinism in \CEU is ``all-inclusive'' and does not depend on additional 
levels of static analysis.
It encompasses the whole language, including memory accesses, system calls, and 
stack-based internal events.
Programmers can always figure out which statement executes next, making runtime 
analysis and debugging easier.
%
\item Static temporal analysis for concurrent statements.
Although execution is deterministic, the compiler of \CEU still advises about 
suspicious statements that can react concurrently to the same event, forcing 
the to serialize them accordingly.
%
\item Safe integration with \C.
\CEU enforces the programmer to deal with pointers representing resources 
through finalization clauses.
When dealing with concurrent system calls, programmers can provide the compiler 
with annotations to remove false positives from the temporal analysis or to 
allow non commutative concurrent behavior.
%
\item First-class timers with dedicated syntax and automatic synchronization.
Given the omnipresence of timers in embedded systems, a dedicated syntax can 
simplify the development and readability of programs.
Furthermore, automatic synchronization releases the programmer from the burden 
of adjusting timers in sequence and in parallel.
%
\end{itemize}

@@

\CEU has a strong influence from Esterel but differ in the most fundamental 
aspect of the notion of time.
%
In \CEU, instead of clock ticks, atomic external event occurrences define time 
units.
%
The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
We consider that this approach is more familiar to programmers and simplifies 
the reasoning about concurrency.
%
As far as we know, \CEU is to encrust this event-triggered notion of time in 
the core of the language, which is a prerequisite for the temporal analysis 
that enables safe shared-memory concurrency.

Another unique aspect of \CEU is the distinction between external and internal 
events.
Internal events support stack-based micro reactions within external reactions, 
providing more fine-grained control for intra-reaction execution.

Like \CEU, many synchronous languages rely on deterministic scheduling to 
preserve intra-reaction determinism (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%
%It is the responsibility of the programmer to specify the execution order for 
%threads, based on either explicit priorities, or source code lexical order 
%(similar to \CEU).
%
\CEU goes one step further and performs a temporal analysis to detect trails 
that, when reordered, change the observable behavior of the program, i.e., 
trails that actually rely on deterministic scheduling.

%
%These languages have a tick-based notion of time similar to Esterel, which 
%prevents the temporal analysis of \CEU.
%
\CEU also guarantees deterministic behavior for timer compositions by adjusting 
inaccuracies in the system clock.

Regarding resource management, Esterel supports a finalization mechanism to 
unconditionally execute a series of statements on abortion.
%
\CEU also tracks pointers representing resources that cross \C boundaries and
enforce the programmer to provide associated finalizers.

@@

We present the design and implementation of \CEU, a synchronous reactive 
language inspired by Esterel with simpler semantics and more fine-grained 
control for intra-reaction execution.

@@

\CEU is a practical language with expressive control constructs, such as 
lexically scoped parallel compositions, convenient first-class timers, and a 
unique stack-based mechanism for internal signalling.
%
Programs interoperate seamlessly with \C, and can take advantage of existing 
libraries, lowering the entry barrier for adoption.
%
\CEU has an open source implementation and bindings for \emph{TinyOS}, 
\emph{Arduino}, and the \emph{SDL} graphical library.%
\footnote{Website of \CEU: \url{http://www.ceu-lang.org/}}

For the past three years, we have been teaching \CEU for undergraduate and 
graduate students in research projects and two hands-on courses on 
\emph{distributed systems} and \emph{reactive programming}.
%, in which the studnets designing protocols, embedded systems, and graphical 
%applications.
%
Our experience shows that students take advantage of the sequential-imperative 
style of \CEU and can implement non-trivial concurrent applications in a few 
weeks.
%
More recently, a company specialized in embedded systems (not related to our 
research group) released a product based on \CEU.
